---
title: "Getting started with STbayes"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with STbayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will walk you through a simple cTADA analysis pipeline using STbayes. We will use some simulated data from a single diffusion trial. First let's load the package and import event and network data. It is possible to include co-variates, but not necessary.

```{r loadpkg, warning=F, message=F}
library(STbayes)
library(ggplot2)
library(tidyr)
library(dplyr)
library(posterior)
```

## Step 1: Importing data

STbayes minimally needs event data and network data. Let's first look at the event data.

```{r import1}
event_data <- STbayes::event_data
head(event_data)
```

Event data gives information  about the spreading of the behavior or information and must contain columns:

- ```id```: Character or numeric individual identities.
- ```trial```: Character or numeric column indicating which trial the event belongs to. There is only one trial in this data-set.
- ```time```: Integer or float indicating when the individual was recorded as first informed/knowledgeable. If an individual **never learned during the observation period, set its value to the duration of the observation period**. These will be treated as right-censored individuals in the likelihood calculation. There are no right censored individuals in this simulated data-set.
- ```t_end```: Integer or float indicating the end of the observation period for each trial, denominated in the same units as time.

Next let's look at the network data:
```{r import2}
edge_list <- STbayes::edge_list
head(edge_list)
```
This edge list gives the network connections of each individual in a long format. You can give a sparse edge list that doesn't include all dyads, but importantly, all individuals must be accounted for in the edge list. The first three columns must be:

- ```trial```: Character or numeric column indicating which trial the networks belong to.
- ```from```: Character or numeric  column of individual identities.
- ```to```: Character or numeric  column of individual identities.

Any remaining columns are descriptively named integer or float edge weights. If you were making a multi-network model, you could add as many columns as you want. Here, there is only one network named ```assoc```.

Let's import the data into a format that can be used with other STbayes functions. I've tried to abstract away as much complexity as possible, but please check the function arguments before sending in the data. The default values might not be what you want! In this case, it's simple---we have no covariates and a single trial with a static network.

```{r import}
data_list <- import_user_STb(event_data, edge_list)
```

This function does quite a bit behind the scenes to process the raw observational data into a named list of variables. I've tried to optimize as much as possible so it shouldn't take more than a second or two unless you're using gigantic networks. Important bits include:

1. *Validates and standardizes inputs.* The function confirms that necessary columns are present, and maps individual and trial identifiers across data frames. The IDs present in the network data are used as the reference for any other data that you supply. For example, if an ID is in the edge list, but not in the event data, it will be automatically added as a censored individual into the event_data. We strongly suggest accounting for all individuals in all data provided to this function to avoid any surprises later on in the analysis.
2. *Preprocesses event data.* The function sorts events by time within each trial and assigns a discrete time index that is used to refer to inter-event intervals. E.g. ```timestep=1``` corresponds to the start of the observation period until the first event. If ```high-res``` mode is used, all durations are set to 1. It also calculates the duration of each of these inter-event intervals. It identifies demonstrators (```time==0```) and right censored individuals (```time>t_end```).
3. *Preprocesses network data.* If static, a single matrix is used for all times. If dynamic, it aligns edge weights across trials and inter-event intervals. If high resolution, it will collapse data into inter-event intervals, multiplying edge weights by transmission weights and then dividing by duration. When fitting, this term will be multiplied out again by duration, so for linear transmission, this is equivalent to running the likelihood in Stan for every single time-step and greatly speeds up model fitting. We note that this pre-processing is not possible complex transmission models.
4. *Preprocesses covariates.* The function recognizes and stores the roles of ILVs as affecting (or not) the intrinsic rate, the social rate or both rates (multiplicative).

## Step 2: Generate a model

Next, we can pass this into a function that automatically creates Stan code that is customised to your data and modelling desires. Each time you generate a model, the function will display the default priors it'll use. You can specify your own priors using the argument ```priors``` and give it a named list.

```{r generate_model}
model_full <- generate_STb_model(data_list, gq = T, est_acqTime = T)
```

This returns a very long string containing the code. The argument ```gq``` indicates whether a generated quantities block will be created. This defaults to TRUE, as it's needed to output the log-likelihood of observations for any kind of model comparison later. The ```est_acqTime``` argument tells STbayes to automatically generate some code to return a posterior distribution of estimated times of events for each individual. 

To quickly check the model code in R, just use ```cat()```. Otherwise, I recommend saving it, as the formatting might need a bit of cleaning before it's readable:

```{r save_model, eval=F}
write(model_full, file="../data/stan_models/my_first_model.stan")
```

If you want to directly modify the code, go ahead. You can supply the file path of the model rather than ```model_obj``` in the next call to ```fit_STb()```. Here, we'll just use the model stored in the variable. 

## Step 3: Fit and save the model

We supply the data_list and model_obj for fitting the model:

```{r fit}
full_fit <- fit_STb(data_list,
    model_full,
    parallel_chains = 4,
    chains = 4,
    cores = 4,
    iter = 4000,
    refresh=1000
)
```

This function calls cmdstanr's $sample(), and you can add any extra valid arguments to pass on here. In this call, I've made sure that we run 4 chains in parallel for 4000 iterations, and only to update us on progress every 1000 iterations. By default, fit_STb() will do iter/2 worth of warmup, and iter/2 worth of sampling. Fitting can take anywhere from seconds to hours depending on your computer and the amount of data. Fitting this dataset _should_ be quick. If it's very slow, you might want to try running this on your friend's computer. I've tried to make things reasonably efficient, but have not parallelized the model block yet.

I strongly recommend using the convenience function below to save the output after fitting. If you only save the fit object to an rda using something like ```save()```, it will not include the data output from chains, making it useless. ```STb_save``` saves these alongside the fit in a single file:

```{r save, eval=F}
STb_save(full_fit, output_dir = "cmdstan_saves", name="my_first_fit")
#> Fit & chains successfully saved 💾 
#> You can load fit again with 👉 readRDS('cmdstan_saves/my_first_fit.rds') 
```

If you don't provide a name, it will use the name of the object.

## Step 4: Viewing model output

We can inspect the parameter estimates using the convenience function below:

```{r summary}
STb_summary(full_fit, digits = 3)
```

The most important output are the intrinsic rate (lambda_0), and the relative strength of social transmission (s), whose interpretations are the same as the NBDA package. The relative strength of social transmission (s = s_prime / lambda_0) is generally what we're after. percent_ST[1] is the estimated percentage of events that occurred through social transmission. The [1] refers to the "assoc" network, as we've only given a single network. If you fit a multi-network model, all networks will have an estimate.  For a number of reasons, STbayes actually fits lambda_0 and social transmission rate (s_prime) on the log scale. The linear transformation of s_prime itself usually isn't reported and is excluded from the output, but you could calculate it yourself from the fit. 

At this point, you are just dealing with a cmdstanr fit, so if you have your own pipeline, adios! However, I have provided a few more useful functions for posterior predictive checks and model comparison.

## Step 5: Posterior predictive checks

STbayes makes it relatively painless to perform posterior predictive checks (PPC) without having to manually simulate data after the fitting process. PPC are important to assess whether or not the model reproduces features of the observed data and make sure that your likelihood reasonably matches the data-generating process. 

Above, we used the argument ```est_acqTime=T```. As the Stan model runs, in each iteration it will generate an estimated event time based on the parameter values for that iteration. With this, we can make two types of PPC.

### Cumulative diffusion curves

We can plot the posterior distribution of cumulative diffusion curves and compare them with the observed curve. First, we create a dataframe for the cumulative number of individuals who experienced an event.
```{r PPC1}
plot_data_obs <- event_data %>%
    filter(time > 0, time <= t_end) %>% # exclude demonstrators (time == 0) and censored (time > t_end)
    group_by(trial) %>%
    arrange(time, .by_group = TRUE) %>%
    mutate(
        cum_prop = row_number() / n(),
        type = "observed"
    ) %>%
    select(trial, time, cum_prop, type) %>%
    ungroup()

# add in 0,0 starting point
plot_data_obs <- bind_rows(
    plot_data_obs,
    plot_data_obs %>%
        distinct(trial) %>%
        mutate(time = 0, cum_prop = 0, type = "observed")
) %>%
    arrange(trial, time)
```

Next, we can extract the draws of estimated acquisition times, excluding warm-up iterations, as a data-frame. We can then pivot it into a long format and calculate the cumulative curves for each draw as we did our observed data. Note that I thinned the draws for the sake of visualization.

```{r PPC2}
draws_df <- as_draws_df(full_fit$draws(variables = "acquisition_time", inc_warmup = FALSE))

# pivot longer
ppc_long <- draws_df %>%
    select(starts_with("acquisition_time[")) %>%
    pivot_longer(
        cols = everything(),
        names_to = c("trial", "ind"),
        names_pattern = "acquisition_time\\[(\\d+),(\\d+)\\]",
        values_to = "time"
    ) %>%
    mutate(
        trial = as.integer(trial),
        ind = as.integer(ind),
        draw = rep(1:(nrow(draws_df)), each = length(unique(.$trial)) * length(unique(.$ind)))
    )

# thin sample for plotting
sample_idx <- sample(c(1:max(ppc_long$draw)), 100)
ppc_long <- ppc_long %>% filter(draw %in% sample_idx)

# build cumulative curves per draw
plot_data_ppc <- ppc_long %>%
    group_by(draw, trial, time) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(draw, trial) %>%
    arrange(time) %>%
    mutate(cum_prop = cumsum(n) / data_list$Q)

# add in 0,0 starting point
plot_data_ppc <- bind_rows(
    plot_data_ppc,
    plot_data_ppc %>%
        distinct(trial, draw) %>%
        mutate(time = 0, cum_prop = 0, type = "ppc")
) %>%
    arrange(trial, time)

# plot it
ggplot() +
    geom_line(data = plot_data_ppc, aes(x = time, y = cum_prop, group = interaction(draw, trial)), alpha = .1) +
    geom_line(data = plot_data_obs, aes(x = time, y = cum_prop), linewidth = 1) +
    labs(x = "Time", y = "Cumulative proportion informed", color = "Trial") +
    theme_minimal()
```

This looks pretty good, the observed curve falls within the variation of the posterior draws. If the observed data fell outside of the draws, you might want to rethink the model specification, whether any influential variables have been excluded, and whether there might be complex transmission processes.

### Estimated vs. observed acquisition times

For each individual, you could also compare the estimated versus observed event times. STbayes' function ```extract_acqTime``` conveniently extracts the times for each individual and re-aligns them with their identities.

```{r PPC3}
acqdata = extract_acqTime(full_fit, data_list)

ggplot(acqdata, aes(x = observed_time, y = median_time)) +
    geom_segment(
        aes(x = observed_time, xend = observed_time, y = median_time, yend = observed_time),
        color = "red",
        alpha = 0.3) +
    geom_point(size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +
    labs(x = "Observed time", y = "Estimated time") +
    theme_minimal()
```

Due to stochastic variation in orders of acquisition between draws, the cumulative curve approach is probably more useful as a PPC.

## Step 6: Model comparison

Often, you'll want to make sure that a model with social transmission is more predictive than a model that excludes it. First, let's refit using a constrained model that only includes an intrinsic rate by using the argument ```model_type="asocial"``` when generating a model. For a fair comparison, make sure the chains and iterations are identical to the full model fit.

```{r model_comparison1}
model_asoc = generate_STb_model(data_list, model_type="asocial")
asocial_fit = fit_STb(data_list,
                      model_asoc,
                      parallel_chains = 4,
                      chains = 4,
                      cores = 4,
                      iter = 4000,
                      refresh=1000)
```

If you are familiar with loo or have your own pipeline, then you can use that to compare these models. STbayes provides a convenient function for performing model comparison. You can provide as many models as you want, here we just compare these two.

```{r model_comparison2}
loo_output = STb_compare(full_fit, asocial_fit, method="loo-psis")
print(loo_output$comparison, simplify = FALSE)
```

If you are not familiar with leave-out-out cross validation, I strongly suggest reading up on it, but the TLDR is that you should pay attention to the elpd_diff and se_diff columns. A model is credibly more predictive if the standard error in differences of ELPD does not cross 0. Here, -21.2 + 5.3 < 0, so we can conclude the full model is more predictive of out-of-sample data than an asocial only model. If you would rather see this graphically:

```{r model_comparison3}
comparison_df <- as.data.frame(loo_output$comparison)
comparison_df$model <- rownames(comparison_df)

ggplot(comparison_df, aes(x = reorder(model, elpd_diff), y = elpd_diff)) +
    geom_point(size = 3) + #elpd_diff
    geom_errorbar(aes(ymin = elpd_diff - se_diff, 
                      ymax = elpd_diff + se_diff), width = 0.2) + #SE of elpd diff
    coord_flip() +
    labs(x = "Model", y = "ELPD Difference", title = "Model Comparison") +
    theme_minimal()
```

PSIS-LOO is an approximation of LOO, and observations with pareto-k diagnostic values >.7 may indicate that the approximation is unreliable. The function above will warn you if that is the case, and you can visually inspect these diagnostics like so:

```{r model_comparison4}
pareto_df = as.data.frame(loo_output$pareto_diagnostics)
ggplot(pareto_df, aes(x=observation, y=pareto_k, color=model))+
    geom_point() +
    scale_color_viridis_d(begin=0.2, end=0.7)+
    geom_hline(yintercept = 0.7, linetype="dashed", color="orange")+
    geom_hline(yintercept = 1, linetype="dashed", color="red")+
    labs(x="Observation", y="Pareto-k value", title="Pareto-k diagnostics")+
    theme_minimal()
```

All of the pareto-k values are below the danger zone.

## Conclusion

These are the basics of using STbayes. You only need event data and network data for a minimal model. However, the package is flexible and you can make use of many other types of data and formulations. Please see the "Advanced recipes" vignette for demonstrations of multi-network models, dynamic network models, and models with ILVs, transmission weights, varying effects, complex transmission, and marginalizing over edge-weight uncertainty.

