---
title: "Model Comparison using Leave Future Out Cross Validation (LFO-CV)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LFO-CV Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(STbayes)
library(igraph)
library(dplyr)
```

Diffusion data is time-series data, and using PSIS-LOO-CV for model comparison (as default in ```STb_compare()```) allows information from future events to influence the predictions for past observations, creating an "overly optimistic" ELPD. To assess the predictiveness of models with time-dependencies, future observations should ideally be excluded as well, in a procedure called leave-future-out cross validation (LFO-CV) ([Bürkner, Gabry & Vehtari 2020](https://www.tandfonline.com/doi/full/10.1080/00949655.2020.1783262)). They propose a method for approximating LFO-CV using pareto-smoothed importance sampling, and only refitting models when the pareto-$k$ value (i.e. variability in importance ratios calculated during the PSIS procedure) exceeds a certain threshold. This greatly reduces the computation time needed to perform LFO-CV, although this will still take more time than using PSIS-LOO-CV or WAIC.

Taking inspiration from code provided by [Bürkner et al.](https://cran.r-project.org/web/packages/loo/vignettes/loo2-lfo.html), I have included functionality for applying approximate LFO-CV to model comparison. This vignette will use simulated data to introduce users to this process.

Let's use the same code from the [Simulating Data for Power Analyses](https://michaelchimento.github.io/STbayes/articles/simulate_data_for_power_analyses.html) vignette to simulate multiple diffusion trials:

```{r}
diffusion_trial <- function(trial_num, obs_duration, g, parameter_values) {
  # Initialize dataframe to store the time of acquisition data
    results <- data.frame(id = 1:N, time = obs_duration + 1, t_end = obs_duration, trial = trial_num)

    # Simulate the diffusion
    for (t in 1:obs_duration) {
        # identify knowledgeable individuals
        informed <- results[results$time <= obs_duration, "id"]

        # identify naive individuals
        potential_learners <- c(1:N)
        potential_learners <- potential_learners[!(potential_learners %in% informed)]

        # break the loop if no one left to learn
        if (length(potential_learners) == 0) {
            results$t_end <- t - 1 # minus 1 so t_end=last learner
            break
        }

        # calculate the hazard for potential learners
        learning_rates <- sapply(potential_learners, function(x) {
            neighbors <- neighbors(g, x)
            C <- sum(neighbors$name %in% informed)
            lambda <- parameter_values[parameter_values$id == x, "lambda_0"] +
                (parameter_values[parameter_values$id == x, "s_prime"] * C)
            return(lambda)
        })

        # convert hazard to probability
        learning_probs <- 1 - exp(-learning_rates)

        # for each potential learner, determine whether they learn the behavior
        learners_this_step <- rbinom(n = length(potential_learners), size = 1, prob = learning_probs)

        # update their time of acquisition
        new_learners <- potential_learners[learners_this_step == 1]
        results[results$id %in% new_learners, "time"] <- t
    }
    return(results)
}

run_sims <- function(num_trials, obs_duration, parameter_values, N, k) {
    # storage for all trials
    all_event_data <- data.frame(
        id = integer(),
        time = numeric(),
        t_end = numeric(),
        trial = integer()
    )
    all_edge_list <- data.frame(
        focal = integer(),
        other = integer(),
        trial = integer(),
        assoc = numeric()
    )
    for (trial in 1:num_trials) {
        # create random regular graph
        g <- sample_k_regular(N, k, directed = FALSE, multiple = FALSE)
        V(g)$name <- 1:N

        results = diffusion_trial(trial, obs_duration, g, parameter_values)

        # create event + network data and append to cumulative dataframes
        event_data <- results %>%
            arrange(time)
        all_event_data <- rbind(all_event_data, event_data)

        edge_list <- as.data.frame(as_edgelist(g))
        names(edge_list) <- c("focal", "other")
        edge_list$trial <- trial
        edge_list$assoc <- 1 # assign e_ij=1 since this is just an edge list
        all_edge_list <- rbind(all_edge_list, edge_list)
        message("Trial", trial, "completed.\n")
    }

    # package up data into list and return it
    final_results <- list(
        "event_data" = all_event_data,
        "networks" = all_edge_list
    )
    return(final_results)
}
```

Let's define our parameters:

```{r}
set.seed(123)

# Parameters
N <- 30 # population size
k <- 5 # degree
log_lambda_0 <- -7
log_sprime <- -5
obs_duration <- 100
num_trials <- 3

#learning parameters
parameter_values <- data.frame(
    id = 1:N,
    lambda_0 = exp(log_lambda_0),
    s_prime = exp(log_sprime)
)

parameter_values$s = parameter_values$s_prime/parameter_values$lambda_0

```

Now, we can run the simulations and fit two models:

```{r}
# run sims
results <- run_sims(num_trials, obs_duration, parameter_values, N, k)

# import data
data_list <- import_user_STb(
    event_data = results[["event_data"]],
    networks = results[["networks"]]
)

# fit full model
model_full <- generate_STb_model(data_list)
fit_full <- fit_STb(data_list, model_full)

# fit constrained model
model_asoc <- generate_STb_model(data_list, model_type = "asocial")
fit_asoc <- fit_STb(data_list, model_asoc)
```

The LFO-CV method presented by Bürkner, Gabry & Vehtari is designed for a single sequential time-series, rather than multiple time-series in the same model. NBDA data may have multiple trials that can be treated as independent diffusions, or at least where the order of the trials is not scientifically meaningful. To accommodate this, I've adjusted the LFO-CV method to proceed in parallel across the trials by default. This approach is appropriate for answering the question: "Given the first i events within each diffusion, how well does the model predict the next M events within each diffusion?" The original approach, sequential LFO-CV, answers the question: "Given all events observed so far, how well does the model predict the next M events (even if those events are in a separate trial at some point in the future)?" This really only makes sense if the ordering of trials is scientifically meaningful, if you expect events from one trial to be predictive of events in another, or if you're reusing individuals between trials and expect that information about them in one trial is meaningful for predicting their behavior in other trials. Also it will take longer to calculate. For the moment, it seems reasonable to me that *parallel is the correct choice for most use-cases of NBDA*, but either way, your choice can be input using argument ```lfo_type``` in the call to ```STb_lfo()```.

In parallel LFO-CV, for each trial, we define a minimum number of observations $L$ that are needed before attempting prediction. We then define $M$ number of observations into the future that we want to predict. STbayes default value is $M=1$. Because we proceed in parallel, if each trial has the same number of individuals, the number of predicted observations at each step is $M*T$. If trials have unequal numbers of individuals, the total number of predicted observations at each step are $M*T_{unfinished}$. I denote $T_{unfinished}$ here because, of course, once we have predicted all observations in a trial, there are none left to be included.

```{r message=FALSE, warning=FALSE, results="hide"}
L <- 5
M <- 1
k_thres = 0.7

output_full <- STb_lfo(
    lfo_type = "parallel", 
    fit = fit_full, 
    STb_data = data_list, 
    stan_model = model_full,
    M=M, L=L,
    k_thres = k_thres
)

output_asoc <- STb_lfo(
    lfo_type = "parallel", 
    fit = fit_asoc, 
    STb_data = data_list, 
    stan_model = model_asoc,
    M=M, L=L,
    k_thres = k_thres
)
```

You can plot the pareto-k values for each observation and see when the model was refitted.

```{r}
plot_ks(attr(output_full, "pointwise")$pareto_k, L=L, k_thres = k_thres)
plot_ks(attr(output_asoc, "pointwise")$pareto_k, L=L, k_thres = k_thres)
```

I've written a convenience function to output a comparison table, but for now I've kept it separate from ```STb_compare()```:

```{r}
lfo_comparison = STb_compare_lfo(output_full, output_asoc)
print(lfo_comparison)
```
We can compare it to the results given by PSIS-LOO. This gives a larger elpd_diff, thus a more optimistic opinion about whether the full model is more predictive.

```{r}
loo_comparison <- STb_compare(fit_full, fit_asoc)
print(loo_comparison$comparison, simplify = F)
```

For illustration, below I calculate the elpd difference by hand. When examining the point-wise ELPD, you'll see that there are a number of NA values. These are the $L$ values we removed before prediction.

```{r}
pointwise_diff = attr(output_full, "pointwise")$elpd_lfo - attr(output_asoc, "pointwise")$elpd_lfo
elpd_diff <- sum(pointwise_diff, na.rm = T)
se_diff <- sqrt(sum(!is.na(attr(output_full, "pointwise")$elpd_lfo))) * stats::sd(pointwise_diff, na.rm = T)
print(paste("elpd_diff:", round(elpd_diff,2), "se_diff", round(se_diff,2)))
```

For the sake of comparison, I rerun with ```lfo_type="sequential"``` below.

```{r message=FALSE, warning=FALSE, results="hide"}
output_full <- STb_lfo(
    lfo_type = "sequential", 
    fit = fit_full, 
    STb_data = data_list, 
    stan_model = model_full,
    M=M, L=L,
    k_thres = k_thres
)

output_asoc <- STb_lfo(
    lfo_type = "sequential", 
    fit = fit_asoc, 
    STb_data = data_list, 
    stan_model = model_asoc,
    M=M, L=L,
    k_thres = k_thres
)
```
```{r}
lfo_comparison = STb_compare_lfo(output_full, output_asoc)
print(lfo_comparison)
```
